---
  title: "Principal Component Analysis: a Simple Example"
---

The data are the results of qualifying examinations for 25 graduate
students in mathematics. diffgeom refers to differential geometry
and reals is for real analysis. Note that the differential geometry
and complex analysis examinations were closed book, while the
remaining three exams were open book.

```{r}
# Load data
setwd("/Users/emmanuel/Dropbox/Work/Teaching/Stanford/Math/Math104/2019-2020/Lectures")
testscores = read.table("testscores.dat.txt")
testscores
```

We now compute the covariance matrix of the data and display it.

```{r}
covX = var(testscores)
covX
```

We compute the correlation matrix and display it as well. The correlation matrix is the same as before, except that the variables are standardized to have unit variance.

```{r}
rcorr = cor(testscores)
rcorr
```

We compute the eigenvalues and eigenvectors of the covariance matrix

```{r}
# Eigen analysis
my.PCs = eigen(covX)
my.PCs
```

We look at the fraction of variance explained by the top principal components

```{r}
cumsum(my.PCs$values)/sum(my.PCs$values)
```

We can of course use R built-in functions

```{r}
pcTest = prcomp(testscores,scale=FALSE)
plot(pcTest)
summary(pcTest)
biplot(pcTest)
```

Normalize variables first

```{r}
pcTest.scaled = prcomp(testscores,scale=TRUE)
summary(pcTest.scaled)
biplot(pcTest)
```
